### 2011-12-09 trunk/papers/smectics/smectics.tex
Next iteration
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1494 40876441-3496-0410-a613-a2fca35b2ae2

\documentclass[8.5pt,twoside,twocolumn]{article}
\oddsidemargin -1.2cm
\evensidemargin -1.2cm
\textwidth 18cm
\headheight 1.0in
\topmargin -3.5cm
\textheight 22cm
\usepackage[super,sort&compress,comma]{natbib} 
\usepackage{mhchem}
\usepackage{times,mathptm}
\usepackage{sectsty}
\usepackage{balance} 
\usepackage{graphicx} %eps figures can be used instead
\usepackage{lastpage}
\usepackage[format=plain,justification=raggedright,singlelinecheck=false,font=small,labelfont=bf,labelsep=space]{caption} 
\usepackage{fancyhdr}
\usepackage{bm}
\usepackage{float}

\pagestyle{fancy}

\newcommand{\update}[1]{{\color{red} #1}}
\newcommand{\e}[1]{\cdot10^{#1}}
\newcommand{\gd}{\dot{\gamma}}

@@

\thispagestyle{plain}
\fancypagestyle{plain}{
\fancyhead[L]{\includegraphics[height=8pt]{LH.pdf}}
\fancyhead[C]{\hspace{-1cm}\includegraphics[height=20pt]{CH.pdf}}
\fancyhead[R]{\includegraphics[height=10pt]{RH.pdf}\vspace{-0.2cm}}
\renewcommand{\headrulewidth}{1pt}}
\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\renewcommand\footnoterule{\vspace*{1pt}% 
\hrule width 3.4in height 0.4pt \vspace*{5pt}} 
\setcounter{secnumdepth}{5}

@@

\fancyfoot{}
\fancyfoot[LO,RE]{\vspace{-7pt}\includegraphics[height=9pt]{LF.pdf}}
\fancyfoot[CO]{\vspace{-7.2pt}\hspace{12.2cm}\includegraphics{RF.pdf}}
\fancyfoot[CE]{\vspace{-7.5pt}\hspace{-13.5cm}\includegraphics{RF.pdf}}
\fancyfoot[RO]{\footnotesize{\sffamily{1--\pageref{LastPage} ~\textbar  \hspace{2pt}\thepage}}}
\fancyfoot[LE]{\footnotesize{\sffamily{\thepage~\textbar\hspace{3.45cm} 1--\pageref{LastPage}}}}
\fancyhead{}
\renewcommand{\headrulewidth}{1pt} 
\renewcommand{\footrulewidth}{1pt}
\setlength{\arrayrulewidth}{1pt}
\setlength{\columnsep}{6.5mm}
\setlength\bibsep{1pt}

\twocolumn[
  \begin{@twocolumnfalse}
\noindent\LARGE{\textbf{Shear Flow of Lamellar Mesophases and Smectic Liquid Crystals}}
\vspace{0.6cm}

\noindent\large{\textbf{O. Henrich\textit{$^{a}$}, K. Stratford\textit{$^{b}$}, D. Marenduzzo\textit{$^{c}$}, P.V. Coveney{$^{a}$} and M.E. Cates \textit{$^{c}$}}
}\vspace{0.5cm}
%Please note that \ast indicates the corresponding author(s) but no footnote text is required.

@@

\footnotetext{\textit{$^{a}$ Centre for Computational Science, University College London, UK}}
\footnotetext{\textit{$^{b}$ Edinburgh Parallel Computing Centre, University of Edinburgh, UK}}
\footnotetext{\textit{$^{c}$ SUPA, School of Physics and Astronomy, University of Edinburgh, UK}}

@@

Instances of soft complex fluids abound. Examples include colloidal
suspensions, where particles or droplets are suspended in a viscous
fluid, living tissues, and liquid crystalline fluids. Well known
examples of colloidal suspensions include paint, cosmetics, and
food products such as ketchup, mayonnaise, and yogurt.
The intracellular network of actin fibres constituting the
cytoskeleton of mammalian cells is a polymeric fluid.
The same is true of the DNA making up the bacterial nucleoid and
eukaryotic chromosomes, and the viral genome of bacteriophages,
where enhanced confinement imparts a strong liquid crystalline
character to the system.
Liquid crystalline fluids themselves, with or without colloidal
inclusions, are of of wide technological importance in devices
such as electronic displays and sensors.

The behaviour under flow, or rheology, of such materials is important
to understand, as it affects many industrial and manufacturing processes.
At the same time, rheological measurements provide a powerful way to
characterise macroscopic properties of a soft material.
The response to applied forces or stresses can be used to quantify,
e.g., the viscosity, or elastic and loss moduli of soft materials.
On shearing, a simple Newtonian fluid exhibits a stress which is
linearly proportional to the applied shear rate; the
constant of proportionality is just the viscosity. In marked
contrast, complex fluids flow is often non-Newtonian and such a simple
linear relation no longer holds. On shearing, the effective viscosity
of a complex viscoelastic fluid may increase or decrease, corresponding
to shear thinkening or shear thinning behaviour. Complex fluids may even
spontaneously phase separate when flowing, giving rise to shear bands.
These are familiar in the rheology of liquid crystals in the vicinity of
the isotropic-nematic transition, and in wormlike micelles.

The rheological properties of a non-Newtonian fluid are typically linked
to the microstructure of the material. Particularly relevant to our
current work is the rheology of layered systems, such as smectics and
cholesterics. Here, experiments may show different rheological results
depending on whether the system has been ordered prior to measurement
(for example, by pre-shearing).
For instance, in the flow of polydomain liquid crystals, which are not
uniformly ordered, the existence of regions with conflicting director
patterns leads to a non-linear relation between shear stress and
applied shear rate.
Another example of strongly non-Newtonian rheology is region I flow
in polymeric nematic liquid crystals. While our understanding of the
physics of region I flow is far from complete, it is thought that a
defect network may form and lead to much enhanced viscosity, with
marked shear thinning. In both these cases, stiffening is usually
associated to the presence of defects in the sample.

Another intriguing type of flow which is possible in layered systems
is permeation,
where individual molecules in the fluid translate while keeping the
defect pattern intact in steady state. This leads to increased dissipation
and hence to a larger viscosity. A well-known example of permeation flow
is found in cholesterics which are sheared or pushed along the direction
of the cholesteric helix. However, one would also expect permeation flows
to be possible when a non-translationally invariant defect pattern is
present in the system; this might be the case in quenched smectics, for
instance. In addition to their non-linear shear-stress curve, lamellar
systems are also of interest for a number of other phenomena.
In block-copolymers one may observe, under some conditions, shear-induced
phase transitions from the microphase separated state to ordered phases
like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}. It would be
desirable to extend our theoretical understanding of such switches.
Further, lyotropic layered liquid crystals have sometimes been observed
to form an onion phase \cite{Panizza96,Iwashita07} which can happen either
spontaneously \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04} or under
shear \cite{Diat93}. The mechanism underlying such pattern formation has
remained elusive to date.

Despite a wealth of experimental results, and the importance of rheology
to many commercial applications, theoretical progress in understanding
flow behaviour of complex fluids has remained a challenge. This is largely
because of the non-linear nature of the equations of motion, and the
coupling of the flow to different microstructures. These either require
drastic approximations or, for practical purposes, resort to computer
simulation. Even here, realistic 3-dimensional calculations are
computationally expensive.
In this work, our focus is the physics of smectics under shear.
While sheared smectics have been studied extensively via simulations
in the literature~\cite{Swift96,Gonnella97,Gonnella98,Xu03,Xu05,Xu06a,Xu06b},
computational constraints have meant almost all these results are limited
to two dimensions. Further, a number of these studies have used solid walls
to drive shear in the hydrodymamics and, inconsistently, sliding periodic
boundaries for the composition variable. Such simulations have a number of
drawbacks: solid walls to drive shear limits the attanable shear rate in
larger systems and so exacerbates finite size effects; mixed boundary
conditions make it difficult to disentangle bulk and boundary effects
in the rheology. In this work, we try to overcome all of these issues.
(One notable example which does tackle a large 3-dimensional simulation
of a complex fluid under shear involved gyroid phases~\cite{Saksena09}.)

We describe smectics via a Brazovskii free energy functional
\cite{Brazovskii75}, which provides a robust framework to study systems
like micelles, surfactants, smectic liquid crystals and block-copolymers.
The governing equations in the model are a convection-diffusion equation
for the scalar order parameter, which represents the density difference
of two fluid components, and the Navier-Stokes equation describing
hydrodynamics. A stress tensor that contains extra contributions owing
to the presence of interfaces between the components is included.
These dynamical equations are solved using a combination of a finite
difference scheme for the convection-diffusion equation, and the
lattice Boltzmann method (LBM)~\cite{Succi} for the Navier-Stokes equation.
We use consistent sliding periodic (Lees-Edwards) boundary conditions for
both convection-diffusion and Navier-Stokes equations. This allows us to
probe the true bulk rheology of the smectic phase for large systems and
high shear rates, as quantified by an appropriate Peclet number.

@@

% KS: this following paragraph doesn't quite make sense to me yet.

Our main finding is that, depending on system parameters, a smectic phase
shows markedly different rheological responses. At lower Peclet
numbers, where thermodynamic effects are relatively strong compared with
flow effects,  shear quickly orders the system into a lamellar phase,
with only a few defects. The rheology in this regime is Newtonian, and
the shear stress is linear in the applied shear rate.
If, on the other hand, thermodynamic forces are larger, we observe that shear actually leads to the nucleation of new defects, and the rheology is highly nonlinear, with the shear rate proportional to $\dot{\gamma}^{\alpha}$, with $\alpha\simeq 0.87$, leading to a shear thinning behaviour. Such a behaviour is qualitatively reminiscent of the one found experimentally in XXX, where the numerical value of $\alpha$ was however different. In this regime we also find that the rheology is history dependent, as the defect density decreases for instance if the shear rate is ramped down gradually. Finally, intriguingly, dimensionality appears to play a key role: within our 3D simulations, we have indeed not been able to enter the nonlinear rheology regime with defect nucleation, and we always observe a shear-induced ordering into a lamellar phase. More research and parameter steering will be needed to make this observation more quantitative.

This paper is organised as follows. In Section II we introduce the model,
including its equilibrium properties, the full stress tensor and the
dynamical equations. We briefly sketch how we derive structural information
from the order parameter and describe our defect quantification procedure,
which measures the local alignment of the lamellar structure.
We present results of the two-dimensional case in in Section III and
discuss the morphology, average and local shear stress densities and
first normal stress differences and their relation to the defect structure.
We also show structure factors for various flow states and flow curves.
We also compare our two-dimensional results with results in three dimensions. 
A summary with conclusions closes the paper in Section V.

\section{Model and Definitions}

The equilibrium phase is found for an order parameter field $\phi({\bm r})\equiv \phi$ that minimises the Landau-Brazovskii free energy functional \cite{Brazovskii75,deGennes}.

\begin{eqnarray}
{\cal F}[\phi]&=&\int d^dx f(\phi({\bm x}))\nonumber\\
&=&\int d^dx\left\{\frac{a}{2}\phi^2+\frac{b}{4}\phi^4+\frac{\kappa}{2} ({\bm \nabla}\phi)^2+\frac{c}{2}(\nabla^2\phi)^2\right\}\label{FE-functional}
\end{eqnarray}

In order to ensure thermodynamic stability the parameters $b$ and $c$ have to be positive.
For $a>0$ there is only one unique solution $\phi\equiv0$ which minimises the functional Eq. \ref{FE-functional} and wich reprsents the the disordered state.
For $a<0$ a microphase separation occurs.
Depending on the sign of $\kappa$ either two homogenous phases coexist ($\kappa>0$) or a transition to an ordered lamellar phase state takes place ($\kappa<0$) as the formation of interfaces become energetically favourable.
The lamellar width is approximately given by $2\pi /\sqrt{|\kappa|/2 c}$, whereas the interfacial width is roughly $\sqrt{2|\kappa|/|a|}$. 
This entails that by varying $a$ (and $b\sim |a|$) by about on decade, the segregation of the two phases varies by about a factor $3$.\\
The dynamics of the binary fluid is governed by an advection-diffusion equation with a diffusive current that is proportional to the chemical potential $\mu$, the functional derivative of ${\cal F}$ with respect to $\phi$.

\begin{equation}\label{OPEOM}
\partial_t \phi +{\bm \nabla}({\bm u}\phi) = M \nabla^2 \frac{\delta \cal F}{\delta \phi}=M \nabla^2 \mu 
\end{equation}

The coefficient $M$ herein plays the role of a mobility coefficient and the diffusivity of the system can be approximated by $D_0\simeq M\;b$.\\
If the fluid is incompressible, the evolution of the velocity ${\bm u}$ obeys the Navier-Stokes equation

\begin{equation}\label{NSE}
\partial_t u_\alpha + (u_\beta \nabla_\beta) u_\alpha = \partial_\beta \sigma_{\alpha \beta}. 
\end{equation}

The stress tensor in this equation reads

\begin{equation}
\sigma_{\alpha \beta}=\eta (\partial_\alpha u_\beta + \partial_\beta u_\alpha)-P^{th}_{\alpha \beta}.
\end{equation}

It consists of a hydrodynamic part, which is proportional to the background viscosity $\eta$ and velocity gradients and a thermodynamical part $P^{th}$ that depends on the order parameter and gradients thereof:

\begin{equation}
P_{\alpha\beta}^{th}=\left(\phi\frac{\delta {\cal F}}{\delta \phi} -f(\phi)\right) \delta_{\alpha\beta} + {\cal D}_{\alpha \beta}(\phi).
\end{equation}

In this last expression the symmetric tensor ${\cal D}_{\alpha\beta})$ has been added to make the pressure tensor divergence free and to guarantee mechanical equilibrium \cite{Evans79}.
The thermodynamic pressure tensor is given by

\begin{equation}
P^{th}_{\alpha \beta}=p\, \delta_{\alpha \beta} + P^{chem}_{\alpha \beta}
\end{equation}

with an isotropic pressure contribution

\begin{eqnarray}
p&=& \frac{a}{2}\phi^2+\frac{3b}{4}\phi^4-\kappa\left(\phi(\nabla^2\phi)+\frac{1}{2}({\bm \nabla}\phi)^2\right)\nonumber\\
& &\quad+c\left(\phi(\nabla^2)^2\phi+\frac{1}{2}(\nabla^2\phi)^2+\partial_{\gamma}\phi\partial_\gamma(\nabla^2\phi)\right)
\end{eqnarray}

and an anisotropic part

\begin{equation}
P^{chem}_{\alpha \beta}=\kappa\partial_\alpha\phi\partial_\beta\phi -c \left(\partial_\alpha\phi\partial_\beta(\nabla^2\phi)+\partial_\beta\phi\partial_\alpha(\nabla^2\phi)\right).
\end{equation}

>>>>

%\documentclass[8.5pt,twoside,twocolumn]{article}
%\oddsidemargin -1.2cm
%\evensidemargin -1.2cm
%\textwidth 18cm
%\headheight 1.0in
%\topmargin -3.5cm
%\textheight 22cm
%\usepackage[super,sort&compress,comma]{natbib} 
%\usepackage{mhchem}
%\usepackage{times,mathptm}
%\usepackage{sectsty}
%\usepackage{balance} 
%\usepackage{graphicx} %eps figures can be used instead
%\usepackage{lastpage}
%\usepackage[format=plain,justification=raggedright,singlelinecheck=false,font=small,labelfont=bf,labelsep=space]{caption} 
%\usepackage{fancyhdr}
%\usepackage{bm}
%\usepackage{float}

%\pagestyle{fancy}
\documentclass[8.5pt,twoside,twocolumn]{article}
\oddsidemargin -1.2cm
\evensidemargin -1.2cm
\textwidth 18cm
\headheight 1.0in
\topmargin -3.5cm
\textheight 22cm
\usepackage[super,sort&compress,comma]{natbib} 
\usepackage{mhchem}
\usepackage{times,mathptm}
\usepackage{sectsty}
\usepackage{balance}

\usepackage{graphicx} %eps figures can be used instead
\usepackage{lastpage}
\usepackage[format=plain,justification=raggedright,singlelinecheck=false,font=small,labelfont=bf,labelsep=space]{caption} 
\usepackage{fancyhdr}
\usepackage{bm}
\pagestyle{fancy}
\usepackage{color}

\newcommand{\update}[1]{{\color{red} #1}}
\newcommand{\e}[1]{\times10^{#1}}
\newcommand{\gd}{\dot{\gamma}}

@@

\thispagestyle{plain}
\fancypagestyle{plain}{
%\fancyhead[L]{\includegraphics[height=8pt]{headers/LH.pdf}}
%\fancyhead[C]{\hspace{-1cm}\includegraphics[height=20pt]{headers/CH.pdf}}
%\fancyhead[R]{\includegraphics[height=10pt]{headers/RH.pdf}\vspace{-0.2cm}}
\renewcommand{\headrulewidth}{1pt}}
\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\renewcommand\footnoterule{\vspace*{1pt}% 
\hrule width 3.4in height 0.4pt \vspace*{5pt}} 
\setcounter{secnumdepth}{5}

@@

\fancyfoot{}
%\fancyfoot[LO,RE]{\vspace{-7pt}\includegraphics[height=9pt]{headers/LF.pdf}}
%\fancyfoot[CO]{\vspace{-7.2pt}\hspace{12.2cm}\includegraphics{headers/RF.pdf}}
%\fancyfoot[CE]{\vspace{-7.5pt}\hspace{-13.5cm}\includegraphics{headers/RF.pdf}}
\fancyfoot[RO]{\footnotesize{\sffamily{1--\pageref{LastPage} ~\textbar  \hspace{2pt}\thepage}}}
\fancyfoot[LE]{\footnotesize{\sffamily{\thepage~\textbar\hspace{3.45cm} 1--\pageref{LastPage}}}}
\fancyhead{}
\renewcommand{\headrulewidth}{1pt} 
\renewcommand{\footrulewidth}{1pt}

\setlength{\arrayrulewidth}{1pt}
\setlength{\columnsep}{6.5mm}
\setlength\bibsep{1pt}

\if{\thispagestyle{plain}
\fancypagestyle{plain}{
%\fancyhead[L]{\includegraphics[height=8pt]{LH.pdf}}
%\fancyhead[C]{\hspace{-1cm}\includegraphics[height=20pt]{CH.pdf}}
%\fancyhead[R]{\includegraphics[height=10pt]{RH.pdf}\vspace{-0.2cm}}
\renewcommand{\headrulewidth}{1pt}}
\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\renewcommand\footnoterule{\vspace*{1pt}% 
\hrule width 3.4in height 0.4pt \vspace*{5pt}} 
\setcounter{secnumdepth}{5}

\fancyfoot{}
\fancyfoot[LO,RE]{\vspace{-7pt}\includegraphics[height=9pt]{LF.pdf}}
\fancyfoot[CO]{\vspace{-7.2pt}\hspace{12.2cm}\includegraphics{RF.pdf}}
\fancyfoot[CE]{\vspace{-7.5pt}\hspace{-13.5cm}\includegraphics{RF.pdf}}
\fancyfoot[RO]{\footnotesize{\sffamily{1--\pageref{LastPage} ~\textbar  \hspace{2pt}\thepage}}}
\fancyfoot[LE]{\footnotesize{\sffamily{\thepage~\textbar\hspace{3.45cm} 1--\pageref{LastPage}}}}
\fancyhead{}
\renewcommand{\headrulewidth}{1pt} 
\renewcommand{\footrulewidth}{1pt}
\setlength{\arrayrulewidth}{1pt}
\setlength{\columnsep}{6.5mm}
\setlength\bibsep{1pt}
}\fi

%\newcommand{\Pe}{\hbox\rm Pe}
\twocolumn[
  \begin{@twocolumnfalse}
\noindent\LARGE{\textbf{Rheology of Lamellar Liquid Crystals in Two and Three Dimensions: A Simulation Study}}
%: A Simulation Study}}
\vspace{0.6cm}

\noindent\large{\textbf{O. Henrich\textit{$^{a,c}$}, K. Stratford\textit{$^{b}$}, D. Marenduzzo\textit{$^{c}$}, P.~V. Coveney{$^{a}$} and M.~E. Cates \textit{$^{c}$}}
}\vspace{0.5cm}
%Please note that \ast indicates the corresponding author(s) but no footnote text is required.

@@

\footnotetext{\textit{$^{a}$ Centre for Computational Science, University College London, 20 Gordon Street, London WC1H 0AJ, UK}}
\footnotetext{\textit{$^{b}$ Edinburgh Parallel Computing Centre, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}}
\footnotetext{\textit{$^{c}$ SUPA, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}}

@@

Instances of soft complex fluids abound in our everyday lives: examples include paint (a colloidal suspension), mayonnaise (an emulsion), and engine oil (a polymer solution). Many soft materials, including the amphiphiles used in skin creams, fabric softeners, and some pharmaceutical products, form lamellar mesophases (also known as smectic liquid crystals) whose perfectly ordered state consists of a regular stack of flat layers. As well as these lyotropic liquid crystals arising in amphiphilic solutions, thermotropic smectics commonly arise in systems of rodlike, mesogenic molecules even without solvent~\cite{deGennes}. Such phases are of wide technological importance in devices
such as electronic displays and sensors.

The flow behaviour (rheology) of smectic liquid crystals affects many industrial and manufacturing processes; conversely, rheological measurements provide important probes of their macroscopic material properties.
For a simple fluid the response to steady applied stress is quantified in the linear (Newtonian) regime by a single viscosity. Lamellar phases are anisotropic, and thus in principle have several viscosities as well as an elastic modulus related to layer compression. However, these linear response parameters become somewhat irrelevant when defects are present: unless their density is very low, the defects soon dominate the stress. Moreover, since the defect pattern is strongly dependent on flow rate, the rheology is generically nonlinear. On shearing, the effective viscosity (ratio of shear stress $\sigma_{xy}$  to strain rate $\dot\gamma$) may either increase or decrease, corresponding
to shear thickening or shear thinning behaviour~\cite{Larson}. Also if a defect texture is present in the system at rest, this may show a viscosity divergence at small $\dot \gamma$ --- i.e., a yield stress.
Finally, in common with other complex fluids, lamellar phases can undergo transitions into heterogeneous states, for instance to form shear bands (layers of unequal $\sigma_{xy}$ or $\dot\gamma$) \cite{Olmsted08}.

Lamellar liquid crystals provide an important paradigm for non-Newtonian fluids in which the microstructure of the material is strongly coupled to the flow. Crucially, the defect texture may depend on past history -- for instance, the strain rate in any prior flow that was suddenly switched off. This rheological hysteresis means that the precise flow protocol, including any pre-shearing applied prior to starting the experiment, needs to be specified. Such hysteresis can arise in polydomain liquid crystals of various types (not just smectics) such as nematics in which the existence and evolution of regions with misaligned directors creates rheological nonlinearity. A closely related example is the so-called `region I flow' in polymeric nematic liquid crystals, where it is thought that a defect network arises, with enhanced viscosity, but with
marked shear thinning as this structure is sheared out at high flow rates \cite{Larson}.

An intriguing complication in addressing the dynamics of layered systems
is permeation,
where individual molecules in the fluid translate while keeping the
layerwise order intact. In steady state, permeation leads to increased dissipation
and hence to a large viscosity in any geometry where flow necessitates such motion \cite{deGennes}. (A well-known example of permeation flow
is found in cholesterics which are sheared or pushed along the direction
of the cholesteric helix.) Permeation is often involved in defect motion such as the growth or shrinkage of incomplete layers at the sites of edge dislocations, and may also be implicated in other forms of topological reconstruction. Such reconstruction arises under shear, sometimes to an extreme degree, for instance in the shear-induced transition from an ordered lamellar stack to an onion phase in lyotropic smectics \cite{Diat93,Panizza96,Iwashita07,Gradzielski03}. (The onion phase comprises a space-filling packing of multilayer vesicles; under some conditions similar defect textures can arise without shear \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04}.) The mechanism underlying such global reconstructions under shear of the layer organization remains unclear. Most theoretical predictions have focussed on buckling and other instabilities of perfectly ordered smectics as the initial trigger for onion formation \cite{Zilman99,Courbin02,Guo2002,Soddemann2004, Stewart2009} but the pathway may also strongly involve the physics of defect motion \cite{Zipfel99, Leon00}. Although we do not see onions in the simulations reported here, in two dimensions we do see a transition from a well ordered phase to a defect-rich phase, whose possible link to onion formation deserves to be explored in future work. Another interesting avenue connects with the observation, in block copolymers and some lyotropic surfactant systems, of a shear-induced
phase transitions from a homogeneous state into ordered phases
like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}. However, theory suggests that thermal noise (causing local fluctuations about the mean state of order) play a central role in this transition~\cite{Cates89}. In the present work we focus on the simplest case in which such thermal fluctuations are ignored, and the time evolution of the microstructure is deterministic (though certainly chaotic and unpredictable
once defects are present).

Despite a wealth of experimental results, and the importance of flow behaviour
to many commercial applications, mechanistic understanding
the rheology of lamellar phases remains limited. This is largely
because of the nonlinear nature of the equations of motion, and the
coupling of the flow to different microstructures. These complications either require
drastic approximations or, as pursued here, use of computer
simulation. However, once defects are present, such simulations are computationally expensive. 
While sheared smectics have been studied extensively via simulations
in the literature~\cite{Swift96,Gonnella97,Gonnella98,Xu03,Xu05,Xu06a,Xu06b},
computational constraints have meant almost all these results are limited
to two dimensions. Even in 2D, the number of layers simulated has often not been large enough to meaningfully explore the dynamics of defect textures which can have a controlling influence on rheology. Moreover, for technical reasons a number of these studies have used boundary conditions (BCs) that are not optimal if one wants to predict the behaviour of bulk samples, for example using solid moving walls to drive the fluid flow rather than periodic BCs. This choice does not minimize finite size effects and, when combined with lattice Boltzmann algorithms, severely limits the shear rates attainable in large systems. Moreover, solid walls for the fluid velocity have sometimes been used in conjunction with periodic BCs for the order parameter \cite{Xu06b} -- a somewhat inconsistent combination that make it difficult to disentangle bulk and boundary effects.

In this work, we try to overcome these technical obstacles, presenting systematic results for large two dimensional simulations with fully periodic boundary conditions.
Guided by those results, we additionally report some very large three dimensional simulations at selected points in the parameter space. These 3D simulations are, we believe, of unprecedented scale for lamellar phases, although even larger ones have recently been used to address the rheology of cubic liquid crystals \cite{Saksena09a, Saksena09b}.

We describe smectics below via a widely-used free energy functional \cite{Gompper, Gonnella97, Xu06b, Cencini07}. This provides a robust generic framework for addressing lamellar ordering without being tied to any specific class of materials (although we do become more specific when estimating suitable parameters for the simulations). The free energy is formulated in terms of a scalar order parameter $\phi({\bf r})$. This can be viewed as describing a composition variable in a binary fluid system stabilized by added amphiphiles; one can then speak of alternating layers of fluid, with interfaces between these layers. A similar interpretation applies to block copolymers where the composition variable describes the relative local density of the two blocks. However, the same order parameter could equally describe the local number density of the midpoints of a rodlike mesogenic molecule (measured relative to the mean) which can develop a periodic density wave that represents a thermotropic smectic state.

The governing equations in our model are a convection-diffusion equation
for the scalar order parameter, and the Navier-Stokes equation describing
the hydrodynamics of fluid flow. The stress tensor for the fluid contains not only a Newtonian viscous term but extra contributions arising from local deviations of the order parameter from its state of least free energy. Diffusive relaxation of the order parameter then causes additional dissipation, resulting in an effective viscosity that can be much higher than the intrinsic Newtonian contribution.
Our dynamical equations are solved using a combination of a finite
difference scheme for the convection-diffusion equation, and the
lattice Boltzmann method (LBM)~\cite{Succi} for the Navier-Stokes equation \cite{Marenduzzo07, Henrich10a,  Henrich10b, Henrich11a,Gonnella97}.
We use sliding periodic (Lees-Edwards) boundary conditions for
both the convection-diffusion and the Navier-Stokes equations. Note that while these boundary conditions impose a fixed integrated strain rate across the whole system they do not impose a uniform strain rate locally; shear bands can therefore arise if the system so chooses. (As it happens, we do not observe shear bands in the range of shear rates we address here.)

%This simulation strategy allows us to probe the bulk rheology of large systems at moderate to high shear rates, as quantified by an appropriate Peclet number, defined below.

@@

%KS: THIS FOLLOWING PARAGRAPH DOESN'T QUITE MAKE SENSE TO ME YET.

Our main finding is that, in two dimensions at least, sheared smectics can show two completely different types of rheological response. At modest rates
of shear, where thermodynamic forces are relatively strong compared with
flow effects,  shear quickly aligns the system into the perfect lamellar phase preferred by thermodynamics,
with only a few defects. The rheology in this regime is quasi-Newtonian, with shear stress linear in the applied shear rate. (There is however a large first normal stress difference.) At higher shear rates, larger thermodynamic forces are generated as layers fail to keep up with the imposed flow. This leads to the nucleation of many new defects, balanced in steady state by annihilation when defects of opposite sign encounter each other. The strain rate $\dot\gamma_c$ needed to enter this nonlinear flow regime depends on thermodynamic and kinetic parameters. A comprehensive exploration of parameter space is not possible, but we give a scaling argument which relates $\dot\gamma_c$ to those parameters through an effective capillary number, formed as the ratio of viscous to thermodynamic stress.
Within the defect proliferation regime, the macroscopic rheology is generically nonlinear, and shows shear thinning. The shear stress appears to vary as a sublinear power law, $\sigma_{xy}\propto\dot{\gamma}^m$, with $m\simeq 0.87$ for typical parameter values, although the dependence of $m$ on the parameters precludes interpretation of this as a universal scaling law.
Such a behaviour is qualitatively reminiscent of the one found experimentally in~\cite{Lu08}, where the numerical value of $m$ was however different. In this strongly non-Newtonian regime we also find that the rheology is history dependent.

Finally, and intriguingly, we find that dimensionality appears to play a key role. Within our 3D simulations, even when performed with parameter values identical to those used in 2D, we have so far not been able to enter the nonlinear flow regime associated with defect proliferation. Instead, we always observe shear-induced ordering into a lamellar phase with quasi-Newtonian rheology. Further research, requiring careful and systematic parameter steering, will be needed to map out quantitatively the onset of the defect proliferation regime, if one exists, in three dimensions.

This paper is organized as follows. In Section \ref{themodel} we introduce the model,
including its equilibrium properties, the full stress tensor and the
dynamical equations. We briefly sketch how we derive structural information
from the order parameter and describe a method for quantitative analysis of the defect density. Section \ref{parametermap} describes the mapping between our simulation parameters and those for physical systems of interest.
We present results for the two-dimensional case in Section \ref{results2D}, discussing
the morphology as a function of flow rate. We also consider average and local shear stress densities and
first normal stress differences, and explore their relation to the defect structure. In addition we report Fourier-space information (structure factors) under flow.
In Section \ref{results3D} we present and compare our results for 3D simulations which, by computational necessity, are less comprehensive than in 2D. 
A summary with conclusions closes the paper in Section \ref{conclusions}.

\section{Model and Definitions} \label{themodel}

The equilibrium phase in a system of volume $V$ is found for an order parameter field $\phi({\bm r})\equiv \phi$ that minimises the Landau-Brazovskii free energy functional \cite{Brazovskii75, Gompper}.
%
\begin{eqnarray}
{\cal F}[\phi] &=& \int d^dx f(\phi,\nabla\phi,\nabla^2\phi) = V\bar f\\ &=& \int d^dx\left\{\frac{a}{2}\phi^2+\frac{b}{4}\phi^4+\frac{\kappa}{2} ({\nabla}\phi)^2+\frac{c}{2}(\nabla^2\phi)^2\right\}\nonumber\label{FE-functional}
\end{eqnarray}
%
In minimizing this free energy at equilibrium, we ignore thermal fluctuations (treated in \cite{Brazovskii75,Cates89}), thereby adopting a mean-field approximation as in previous literature \cite{Kendon01,Gonnella97, Gonnella98, Xu03, Xu05, Xu06a, Xu06b}. This should be a good approximation far from the ordering transition at which $\phi$ first becomes nonuniform. The uniform state has $\phi = 0$ and hence ${\cal F}= 0$.

The parameter range relevant to lamellar phases has $\kappa < 0$ and $c>0$. Then, if we suppress the quartic term, the free energy density of a sinusoidal lamellar state $\phi({\bf r}) = \phi_k\cos({\bf k}.{\bf r})$ obeys $\bar f = (a+\kappa k^2+ c k^4)\phi_k^2/4$. This is minimized by choosing $k=q\equiv\sqrt{|\kappa|/2c}$; the free energy density is then $\bar f = (a-\kappa^2/4c)\phi_q^2/4$ so that ordering will occur for $a<a_c=\kappa^2/4c$. The amplitude $\phi_q$ is then unbounded without the (positive) quartic term, since $\bar f$ can in that manner become infinitely negative. Restoring the quartic term now causes saturation at $\phi_q\sim \sqrt{4(a_c-a)/3b}$, creating also nonlinear harmonics in the density profile. The final free energy density obeys $\bar f \simeq (a_c-a)^2/b$. By similar arguments, the characteristic elastic modulus associated with microstructural deformations is
%
\begin{equation}
\bar g\sim \frac{(a_c-a)a_c}{4b}
\label{barg}
\end{equation}
%
which is the energy density of the gradient terms only. The terms involving $\phi^2$ and $\phi^4$ in Eq.~\ref{FE-functional} are unaltered by affine deformations which conserve density locally; hence those terms do not contribute to $\bar g$.

The dynamics of the order parameter is governed by an advection-diffusion equation with a diffusive current of $\phi$ dependent on the chemical potential $\mu$, the functional derivative of ${\cal F}$ with respect to $\phi$, as follows:
%
\begin{equation}\label{OPEOM}
\partial_t \phi +{\nabla}({\bm u}\phi) = M \nabla^2 \frac{\delta \cal F}{\delta \phi}=M \nabla^2 \mu 
\end{equation}
%
Here $M$ is a mobility coefficient. A wavevector-dependent linearized collective diffusivity $D_k$ can then be defined via $\dot\phi_k = -k^2D_k\phi_k$, whose value at the ordering wavevector obeys $D_q = M(a-a_c)/2$ (this is negative, signifying the instability). In contrast, an estimate of the molecular diffusivity can be made by considering zero wavevector, setting $b = -a$ (see below) and examining the relaxation equation for small deviations around the uniformly ordered state at $\phi = \pm 1$. This gives $\dot\phi = 2bM\nabla^2\phi$ so that the molecular diffusivity is of order $\tilde D = 2Mb$.
